Litrepl: Literate Paper Processor Promoting Transparency More Than Reproducibility

### 2025-01-01 paper/paper.tex
Update paper

\newcommand{\es}{\begin{enumerate}\item}
\newcommand{\ei}{\item}
\newcommand{\ee}{\end{enumerate}}

@@

Litrepl is a simple and lightweight text processing tool designed to recognize
and evaluate code sections within Markdown or LaTeX documents. This
functionality is useful for both batch document section evaluation and
interactive coding within a text editor, provided a straightforward integration
is established.

@@

Concurrently, another concept of human-computer interaction called the
Read-Evaluate-Print Loop (REPL) gained traction, notably within the LISP
community and through the APL language designed for mathematical computations.
The combination of a command-line interface and a language interpreter enables
incremental and interactive programming, allowing users to directly modify the
interpreter state. By maintaining human involvement in the loop, this approach
facilitates human thought processes \cite{Spence1975apl, McCarthy1959recfun,
Iverson1962apl}.

A significant milestone in this area was the development of the IPython project,
which later evolved into the Jupyter Project. The creators introduced a new
document format called the Notebook, characterized by a series of logical
sections of various types, including text and code, which could directly
interact with programming language interpreters. This interactive communication,
akin to REPL style programming, allows the creation of well-structured documents
suitable for presentations and sharing. The concept underpinning these
developments is termed Literate Computing, which includes goals of spanning a
wide audience range, boosting reproducibility, and fostering collaboration. To
achieve these objectives, several technical decisions were made, notably the
introduction of bidirectional communication—between the computational core,
known as the Jupyter Kernel, and the Notebook acting as a client, along with
another layer of client-server communication between the Notebook web server and
the user’s web browser \cite{Perez2007IPython, Granger2021litcomp,
Kluyver2016jupnb}.

4. While we recognize the importance of various goals within the Literate
Computing framework, we argue that reproducibility is paramount. Successfully
addressing reproducibility would alone suffice to enhance communication over
time among research project participants and significantly expand the audience.
However, as it became clear (\cite{Dolstra2010, Vallet2022}), this challenge
extends far beyond the scope of a human-computer interaction system, and even
beyond the typical boundaries of software distribution management for a
particular programming language. A comprehensive solution to the software
deployment problem should operate at the entire operating system level.

5. Consequently, we propose focusing on transparency in human-computer
interaction rather than on reproducibility. We present "Litrepl," a tool
designed to integrate the REPL interactive programming style into existing
editors in a straightforward yet potent manner.

6. First, we utilize simple bidirectional text streams for inter-process
communication with language interpreters to evaluate code. Second, we advocate
for the reuse of existing text document formats. In both the Markdown and LaTeX
formats that we have implemented, simplified parsers are utilized to distinguish
code and result sections from the rest of the document. As of now, we support
Python and Shell interpreter families, as well as a custom AI communication
interpreter. Finally, we strive to leverage POSIX \cite{POSIX2024} operating
system facilities as much as possible.

@@

This document contains a Python code section and an empty result section marked
with the corresponding Latex environment tags. To "execute" the document we
pipe it though the Litrepl processor as follows:

@@

%% ignore %%
\begin{figure*}[hbt!]
  \centering
  \begin{minipage}{\textwidth}
  \begin{verbatim}
  document       ::= (code | result | ignore | text)*
  code           ::= (code-normal | code-comment)
  result         ::= (result-normal | result-comment)
  code-normal    ::= "\begin{MARKER}" text "\end{MARKER}"
  code-comment   ::= "% MARKER" text "% noMARKER"
  result-normal  ::= "\begin{result}" text "\end{result}"
  result-comment ::= "% result" text "% noresult"
  ignore         ::= "% ignore" text "% noignore"
  text           ::= ...
  \end{verbatim}
  \end{minipage}
  \caption{An illustrative grammar template for LaTeX documents where marker
  serves as a parameter configured via command-line arguments for each supported
  interpreter class.}
  \label{fig:document-representation}
\end{figure*}
%% noignore %%

@@

Template grammars similar to the illustrative example in Figure
\ref{fig:document-representation} are encoded for Markdown and LaTeX formats.
Before each run, Litrepl calls Lark\cite{Lark} to compile a customized parser
and uses it to access the sections.

@@

The restriction to text-only data types presents a more fundamental limitation.
Litrepl lifts this restriction by supporting text-only document formats. Both
LaTeX and Markdown incorporate rich data without encoding it directly, instead
relying on side channels, such as the file system or network, to refer to such
objects. Consequently, Litrepl shares the advantages of good representation in
well-known version control systems, and the disadvantages, such as the need to
explicitly organize side-channel data transfer.

>>>>

\newcommand{\es}{\begin{enumerate}\item}
\newcommand{\ei}{\item}
\newcommand{\ee}{\end{enumerate}}
\newcommand{\Latex}{\LaTeX\ }

@@

Litrepl is a simple and lightweight text processing tool designed to recognize
and evaluate code sections within Markdown or \Latex documents. This
functionality is useful for both batch document section evaluation and
interactive coding within a text editor, provided a straightforward integration
is established.

@@

Concurrently, a concept of human-computer interaction called the
Read-Evaluate-Print Loop or ''REPL'' gained traction, notably within the LISP
an APL communities.\cite{Spence1975apl, McCarthy1959recfun, Iverson1962apl}

The combination of a command-line interface and a language interpreter enables
incremental and interactive programming, allowing users to directly modify the
interpreter state. By maintaining human involvement in the loop, this approach
is believed to facilitate human thought processes.

A significant milestone in this area became the IPython terminal, which later
evolved into the Jupyter Project. The creators introduced a new document format
called the Notebook, characterized by a series of logical sections of various
types, including text and code, which could directly interact with programming
language interpreters. This interactive communication, akin to REPL style
programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
Literate Computing, which includes goals of spanning a wide audience range,
boosting reproducibility, and fostering collaboration. To achieve these
objectives, several technical decisions were made, notably the introduction of
bidirectional communication—between the computational core, known as the Jupyter
Kernel, and the Notebook acting as a client, along with another layer of
client-server communication between the Notebook web server and the user’s web
browser \cite{Perez2007IPython, Granger2021litcomp, Kluyver2016jupnb}.

While we recognize the importance of these goals within the Literate Computing
framework, we argue that reproducibility is paramount. Successfully addressing
reproducibility would alone suffice to enhance communication over time among
research project participants and significantly expand the audience.  However,
as it became clear (\cite{Dolstra2010, Vallet2022}), this challenge extends far
beyond the scope of a human-computer interaction system, and even beyond the
typical boundaries of software distribution management for a particular
programming language. A comprehensive solution to the software deployment
problem should operate at the entire operating system level.

After understanding the scale and complexity of the reproducibility problem, we
suggest to change the focus of human-computer interaction towards simplicity and
transparency. We saw an opportunity to implement a tool named "Litrepl", which
would offer the same familiar REPL-style editing, be compatible with existing
text editors, thus maintaining familiar editing practices, and have a gradually
smaller codebase.

We accomplish this through the following steps. First, we utilize simple
bidirectional text streams for inter-process communication with language
interpreters to evaluate code. Second, we advocate for the reuse of existing
text document formats. In both the Markdown and \Latex evaluators that we have
implemented, simplified parsers are utilized to distinguish code and result
sections from the rest of the document. As of now, we support Python and Shell
interpreter families, as well as a custom AI communication interpreter. Finally,
we strive to leverage POSIX \cite{POSIX2024} operating system facilities as much
as possible.

@@

This document contains a Python code section and an empty result section marked
with the corresponding \Latex environment tags. To "execute" the document we
pipe it though the Litrepl processor as follows:

@@

%% ignore %%
\begin{figure*}[hbt!]
  \centering
  \begin{minipage}{\textwidth}
  \begin{verbatim}
  document       ::= (code | result | ignore | text)*
  code           ::= (code-normal | code-comment)
  result         ::= (result-normal | result-comment)
  code-normal    ::= "\begin{MARKER}" text "\end{MARKER}"
  code-comment   ::= "% MARKER" text "% noMARKER"
  result-normal  ::= "\begin{result}" text "\end{result}"
  result-comment ::= "% result" text "% noresult"
  ignore         ::= "% ignore" text "% noignore"
  text           ::= ...
  \end{verbatim}
  \end{minipage}
  \caption{An illustrative grammar template for \Latex documents where marker
  serves as a parameter configured via command-line arguments for each supported
  interpreter class.}
  \label{fig:document-representation}
\end{figure*}
%% noignore %%

@@

Template grammars similar to the illustrative example in Figure
\ref{fig:document-representation} are encoded for Markdown and \Latex formats.
Before each run, Litrepl calls Lark\cite{Lark} to compile a customized parser
and uses it to access the sections.

@@

The restriction to text-only data types presents a more fundamental limitation.
Litrepl lifts this restriction by supporting text-only document formats. Both
\Latex and Markdown incorporate rich data without encoding it directly, instead
relying on side channels, such as the file system or network, to refer to such
objects. Consequently, Litrepl shares the advantages of good representation in
well-known version control systems, and the disadvantages, such as the need to
explicitly organize side-channel data transfer.
